Human capital mediates natural selection in contemporary humans

### 2022-03-19 why-natural-selection.Rmd
Work on graphs. Use medians not means for percentage differences.

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

@@

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))

}

@@

pval_geom_point <- function (size = 1.6, stroke = 0.8, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

  if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

add_n <- function (var, famhist_var = var, dv = "RLRS", data = famhist) {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv, data)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

As yet we know little about the social mechanisms behind natural selection. A
potential explanation comes from the economic theory of fertility [@becker1960economic]. In this theory, higher potential earnings
have two opposite effects on fertility: a fertility-increasing *income effect*
(higher income makes children more affordable), and a fertility-lowering
*substitution effect* (time spent on childrearing has a higher cost in foregone
earnings). Thus, an individual's *human capital* -- skills and personality
traits which are valuable in labour markets -- can increase or decrease his or
her fertility. This can lead to selection on inherited variants which are linked
to human capital. Furthermore, the economic theory predicts systematic variation
in the relative strength of the income and substitution effects across different
social groups.

@@

We then show that a simple model of human capital, education and fertility
choices can give rise to these empirical results. To test the theory, we show
that polygenic scores' correlation with fertility is associated with their
correlation with education and earnings. Furthermore, a mediation analysis shows
that part of the correlation with fertility is indeed *via* education. Thus, contemporary natural selection on polygenic scores can be explained by scores' correlation with earnings-increasing human capital.

Lastly, we discuss the effects of natural selection. While measured effects on
mean genetic scores are small, natural selection may substantially increase the
correlation between genetic scores and income, increasing genetic differences 
between different social groups, and thus making the "genetic lottery"
[@harden2021genetic] more unfair.

@@

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias within the sample [@10.1093/aje/kwx246].

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.2) +
      facet_wrap(~score_name, ncol = 6) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "square")) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test more directly for natural selection, we regress respondents' relative 
lifetime reproductive success ($RLRS_i$) on their polygenic scores (PGS):

@@

RLRS is defined as respondents' number of children, divided by the mean number
of children of people born in the same year. The "selection effect", $\beta$,
reflects the strength of natural selection within the sample. In fact, since
polygenic scores are normalized, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic].[^normalization] Note that
equation \eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a correlational phenomenon: polygenic scores which correlate with
high fertility are being selected for, whatever the underlying causal
relationship.

@@

drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)
drake::loadd(res_sex)

res_wt_combined <- bind_rows(
  None                = res_all %>% 
                          filter(
                            reg.type == "controlled", 
                            dep.var == "RLRS"
                          ) %>% 
                          select(-reg.type, -dep.var, -score_name),
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

res_significance <- res_wt_combined %>% 
                      group_by(term) %>% 
                      summarize(
                        all = all(p.value < 0.05/33), 
                        any = any(p.value < 0.05/33)
                      )
res_sig_all <- sum(res_significance$all)
res_sig_any <- sum(res_significance$any)

@@

rel_sizes <- res_wt_cb_wide %>% 
                filter(
                  consistent, 
                  p.value < 0.05/length(score_names)
                ) %>% 
                mutate(
                  Geographical = Geographical/None,
                  `Age/Qualification` = `Age/Qualification`/None,
                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
                ) %>% 
                select(-consistent, -None, -p.value, -estimate_females) %>% 
                mutate(
                  term = pretty_names(term)
                )

@@

To correct for ascertainment bias, we weight participants using population data.
We try three alternative weighting schemes: (1) "Geographical" weighting, by
geography, age and presence/absence of a partner; (2) age and highest
educational qualification; and (3) for women only, age, highest qualification,
and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
selection effects among the entire sample, estimated with the three weighting
schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
scores are significant at $p < 0.05/33$ under all four weighting schemes, with
`r res_sig_any` scores reaching significance under at least one scheme.[^balance-diverse-def]
Weighting makes a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic
variables.[^weighting]

@@

[^weighting]: To make interpretation straightforward, we do not use weights in 
results below unless noted. However, all our qualitative results are robust to
using weights.

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income and education, and
family structure variables including age at first live birth, presence of a
partner, and lifetime number of sexual partners. Figure
\@ref(fig:plot-income-educ-level) plots selection effects for each polygenic
score, grouping respondents by age of completing full-time education, and by
household income. Effects are larger and more significant for the lowest income
category, and for the lowest education category. These results are robust to
controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

Turning to family structure, we split respondents by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
Selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

@@

n_regs <- as.double(nrow(res_partners_joint))
res_partners_joint %>% 
      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
                        score_col = score_name)

@@

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      standard_ggplot(fill_col = Household, n_regs = n_regs, 
                        score_col = score_name)
```

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
separately for each tercile of AFLB. Effects are strikingly different
across terciles. Educational attainment, ADHD and MDD are selected for amongst
the youngest third of mothers, but selected against among the oldest two-thirds.
Similarly, several polygenic scores for body measurements are selected against
only among older mothers. The correlation between effect sizes for the youngest
and oldest terciles is `r cor_age_flb`.

@@

To investigate this further, we estimate equation \@ref(eq:regression) among
females, *controlling* for AFLB. In `r opp_signed` out of 
`r length(score_names)` cases, effects change sign when controls are added. The
correlation between effect sizes controlling for AFLB, and raw effect sizes, is
`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: a correlation with AFLB, and an opposite-signed correlation with 
number of children after AFLB is controlled for.

@@

We emphasize that our categories are not necessarily exogenous to the polygenic
scores. For example, in the data, and also in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related
polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
differences in equation \@ref(eq:regression) across subgroups constrain the set
of possible explanations for natural selection. As we show below, a simple
economic model predicts the differences we find.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 9 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise. There is also a literature showing negative correlations
between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

@@

res_mediation %<>% mutate(
                     # 0.025 not 0.05 because two-sided
                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/33),
                     pos_sig = sig & prop_ind > 0
                   )

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with the economic theory of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

  if (conf_int) {
    confint_scale <- scale_linetype_manual(
                       name   = "", 
                       values = c("95% c.i. uncorrected" = 1),
                       guide  = guide_legend(order = 1)
                     )
    confint_geom_segment <- confint_geom_segment()
  } else {
    confint_scale <- NULL
    confint_geom_segment <- NULL
  }

@@

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 0,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

@@

pval_geom_point <- function (size = 1.6, stroke = 0.9, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey55", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

  scale <- if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale <- scale_color_brewer(aesthetics = aesthetics, ...)
    shades::lightness(scale, shades::scalefac(0.75))
  }

  scale
}

@@

add_n <- function (var, famhist_var = var, dv = "RLRS", data = famhist, 
                     reverse = FALSE) {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv, data)
  # this is a horrible hack which just happens to work
  # check all Ns manually!
  if (reverse) n <- rev(n)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

As yet we know little about the social mechanisms behind natural selection. The
economic theory of fertility [@becker1960economic] offers a potential
explanation. Higher potential earnings have two opposite effects on fertility: a
fertility-increasing *income effect* (higher income makes children more
affordable), and a fertility-lowering *substitution effect* (time spent on
childrearing has a higher cost in foregone earnings). Thus, an individual's
*human capital* -- skills and personality traits which are valuable in labour
markets -- can increase or decrease their fertility. Genetic variants  which are 
linked to human capital will then be selected for or against. Also, the economic
theory predicts that the relative strength of income and substitution effects
will vary systematically across different social groups.

@@

We then show that a simple model of human capital, education and fertility
choices can give rise to these empirical results. At higher incomes, the income
and substitution effects are balanced, while among lower-income people, or
single parents who face a bigger time burden from childcare, the substitution
effect dominates. The theory predicts that polygenic scores' correlation
with fertility is associated with their correlation with education and earnings,
and we confirm this. We then run a mediation analysis, which shows that part of the
correlation with fertility is indeed mediated by educational attainment. 
Thus, contemporary natural selection on polygenic scores can be explained 
by scores' correlation with earnings-increasing human capital.

Lastly, we discuss the effects of natural selection. While our estimated effects 
on measured polygenic scores are small, natural selection substantially 
increases the correlation between polygenic scores and income, increasing 
genetic differences between different social groups, and thus making the 
"genetic lottery" [@harden2021genetic] more unfair.

@@

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias in the sample [@10.1093/aje/kwx246].

```{r plot-means-over-time, fig.height = 7, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Symbols show means for 5-year intervals. Bars are 95% confidence intervals. Triangles denote a significant linear increase or decrease over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey70", 
                       size = 3) +
      geom_point(aes(y = score), size = 1.5) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c(
                                      "-" = "triangle down open", 
                                      "+" = "triangle open", 
                                      "o" = "circle open"
                                    )) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test for natural selection more directly, we regress respondents' relative 
lifetime reproductive success (RLRS) on each polygenic score (PGS):

@@

RLRS is defined as respondent $i$'s number of children, divided by the mean number
of children of people born in the same year. The "selection effect", $\beta$,
reflects the strength of natural selection within the sample. In fact, since
polygenic scores are normalized, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic].[^normalization] Note that
equation \eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a matter of correlation not causation: polygenic scores which 
correlate with high fertility are being selected for, whatever the underlying 
causal mechanism.

@@

drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_wt_van_alten_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_unweighted)
drake::loadd(res_sex)

res_wt_combined <- bind_rows(
  None                = res_unweighted %>% 
                          select(-score_name) %>% 
                          filter(term != "(Intercept)"),
  Main                = res_wt_van_alten_weights,
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

n_sig <- sum(res_wt_van_alten_weights$p.value < 0.05/33)

@@

rel_sizes <- res_wt_cb_wide %>% 
                filter(
                  consistent, 
                  p.value < 0.05/length(score_names)
                ) %>% 
                mutate(
                  Geographical        = Geographical/None,
                  `Age/Qualification` = `Age/Qualification`/None,
                  Main                = Main/None,
                  `Age/Qual/AFLB`     = `Age/Qual/AFLB`/estimate_females
                ) %>% 
                select(-consistent, -None, -p.value, -estimate_females) %>% 
                mutate(
                  term = pretty_names(term)
                )

@@

Figure \@ref(fig:plot-res-wt) plots selection effects in the whole
sample.[^balance-diverse-def] To correct for ascertainment bias, we use
participant weights from @van2021effects, which match the UK Biobank eligible
population on sex, birth year, location, education, employment, health,
household size and tenure, number of cars and age at death. Weighting makes a
large difference: effect sizes go up by a mean of 
`r scales::percent(mean_rel$Main - 1)`.[^weighting] `r n_sig` out of 33 weighted
selection effects are significant at $p$ < 0.05/33.

@@

[^weighting]: We use these weights throughout. All our qualitative results are
robust if we run unweighted regressions. Appendix Table \@ref(tab:tbl-res-wt)
shows results from alternative weighting schemes.

@@

res_wt_combined %>% 
      filter(Weights %in% c("Main", "None")) %>% 
      mutate(
        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      standard_ggplot(fill_col = Weights, n_regs = n_regs)

@@

We now show the empirical puzzles which motivate our economic model. Each
concerns differences in the strength of natural selection across different
subgroups in the sample. We re-estimate \@ref(eq:regression) splitting the
sample by demographic and social variables, including income and education, and
family structure variables including age at first live birth, presence of a
partner, and lifetime number of sexual partners.

```{r calc-pct-diffs}
drake::loadd(res_edu)
drake::loadd(res_income)
drake::loadd(res_partners_joint)
drake::loadd(res_with_partner)
res_partners_joint %<>% filter(grepl(":", term))
res_with_partner %<>% filter(grepl(":", term))

res_edu_wide <- full_join(
                  res_edu %>% filter(age_fte_cat == "< 16"),
                  res_edu %>% filter(age_fte_cat == "> 18"),
                  by = "score_name"
                )

res_income_wide <- full_join(
                     res_income %>% filter(income_cat == 1),
                     res_income %>% filter(income_cat == 5),
                     by = "score_name"
                   )

res_partners_wide <- full_join(
                       res_partners_joint %>% filter(grepl("lo_partnersFALSE", term)),
                       res_partners_joint %>% filter(grepl("lo_partnersTRUE", term)),
                       by = "score_name"
                     )

res_with_partner_wide <- full_join(
                           res_with_partner %>% 
                             filter(grepl("with_partnerFALSE", term)),
                           res_with_partner %>% 
                             filter(grepl("with_partnerTRUE", term)),
                           by = "score_name"
                         )

median_pct_diff <- function (dfr) {
   prop <- dfr %>% filter(
                    p.value.x < 0.05/33,
                    sign(estimate.x) == sign(estimate.y)
                  ) %>% 
                  mutate(
                    pct_diff = estimate.x/estimate.y
                  ) %>% 
                  pull(pct_diff) %>% 
                  mean()
   # percentage difference:
   prop - 1
}

pct_diff_edu <- median_pct_diff(res_edu_wide)
pct_diff_income <- median_pct_diff(res_income_wide)
pct_diff_n_partners <- median_pct_diff(res_partners_wide)
pct_diff_with_partner <- median_pct_diff(res_with_partner_wide)
```

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
education category, and for the lowest income category. The median percentage
difference between the lowest and highest education categories, among scores
which are significant for the lowest category and have the same sign across
categories, is `r scales::percent(pct_diff_edu)`. Between the lowest and
highest income categories, it is `r scales::percent(pct_diff_income)`. These
results are robust to controlling for respondents' age (Appendix section
\@ref(sec:age-control)). Turning to family structure, we split respondents by
lifetime number of sexual partners, at the median value of 3 (Figure
\@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
significant among those with more than 3 lifetime partners, with a median percentage
difference of `r scales::percent(pct_diff_n_partners)`. Next we split
respondents by whether they were living with a spouse or partner at the time of
interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The median percentage difference is 
`r scales::percent(pct_diff_with_partner)`.[^sex]

@@

@@

n_regs <- as.double(nrow(res_partners_joint))
res_partners_joint %>% 
      mutate(
        `N partners` = add_n(`N partners`, "n_partners_split", reverse = TRUE,
                               data = famhist %>%
                               mutate(n_partners_split = n_partners <= 3))
      ) %>% 
      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
                        score_col = score_name)

@@

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        Household = add_n(Household, "with_partner", reverse = TRUE)
      ) %>% 
      standard_ggplot(fill_col = Household, n_regs = n_regs, 
                        score_col = score_name)
```

@@

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
separately for each tercile of AFLB. Effects are strikingly different
across terciles. Educational attainment, ADHD and MDD are selected for amongst
the youngest third of mothers, but selected against among the oldest two-thirds.
Similarly, several polygenic scores for body measurements are selected against
only among older mothers. The correlation between effect sizes for the youngest
and oldest terciles is `r cor_age_flb`. To investigate this further, we estimate
equation \@ref(eq:regression) among females, *controlling* for AFLB. In 
`r opp_signed` out of `r length(score_names)` cases, effects change sign when
controls are added. The correlation between effect sizes controlling for AFLB,
and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
through two opposing channels: a correlation with AFLB, and an opposite-signed
correlation with number of children after AFLB is controlled for.

@@

We emphasize that these categories are not exogenous to polygenic scores. For
example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
our theoretical model -- education and age at first live birth are choice
variables, which are endogenous to a person's human capital and to relevant
polygenic scores. Nevertheless, differences in selection effects across
subgroups constrain the set of possible explanations. A good theory of contemporary
natural selection needs to show how these differences come about. As we describe
below, a model based on the economic theory of fertility can do just that.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 8 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
AFLB and socio-economic status in the parents' generation, using the 1971 
Townsend deprivation score of respondents' birthplace as a proxy for income
[@townsend1987deprivation]. Results show the same pattern as for the
respondents' generation. Effect sizes are larger and more often significant in
the most deprived areas (Appendix Figure \@ref(fig:plot-siblings-townsend)).
Effects are larger among younger fathers and mothers, and change sign when
controlling for AFLB (Appendix Figures \@ref(fig:plot-age-birth-parents-cross),
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a "quantity-quality
tradeoff" between parents' number of children and number of grandchildren. We
don't find any: in fact, the correlation between respondents' and parents' RLRS
is positive ($\rho$ = `r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).

@@

Thus, a simple economic model can explain many of our results. Other empirical
work in economics also supports the link from human capital to fertility.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted
by @becker1981treatise. A related literature shows negative correlations between 
IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

@@

res_mediation %<>% mutate(
                     # 0.025 not 0.05 because two-sided
                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/
                                                            nrow(res_mediation)),
                     pos_sig = sig & prop_ind > 0
                   )
